Landau quantization of nearly degenerate bands

### 2018-09-16 paper.tex
Updates from ShareLaTeX

For simplicity, we assume that the Pauli matrices in \q{eq:Rashba-Hamiltonian} correspond to spin: $s_i{=}\hbar \sigma_i/2$. Let us evaluate the effective Hamiltonian $\calh(\bk)$  along the cyclotron trajectory $\frako_0$ with radius $k$. In the  energy basis of the Rashba Hamiltonian $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], 
\e{\calh(\bk)=-\alpha k\tau_z{+}\f{g_s}{2}\mu_{\sma{B}}B\tau_x \red{-} \f{eB\hbar}{2m} (\tau_x-I) \epsilon_{\alpha\beta} k_{\red{\alpha}}\nabla_{\red{\beta}}\theta,\la{rashbaeffham}}
where $\tau_i$ are a new set of Pauli matrices with $\tau_z{=}{+} 1$ and ${-}1$ corresponding to the outer and inner orbits, respectively \red{[I switched the sign in the equation above to emphasize the sign difference.]}. $e^{i\theta_{\bk}}{=}(ik_x{-}k_y)/k$ is the angle of the in-plane spin-orbit field $({-}\alpha k_y,\alpha k_x,0)$, which winds once over $\frako_0$; the spin of our basis vectors likewise winds, in parallel (or anti-parallel) alignment with this field (see Fig. (\ref{fig:orbits})(b)). Terms proportional to the rate of spin rotation  ($\nabk \theta$) originate from the  non-abelian Berry connection. The first term of  $\calh$ corresponds to the spin-orbit-induced splitting of the energy-degenerate bands;
the remaining terms may be interpreted as B-induced deformations of the bands.   It is well-known that the Zeeman term tends to polarize spin in the direction ($\vec{z}$) of $\bB$; remarkably, the non-abelian Berry term   also has a polarizing effect, but with opposite sign. These two effects completely cancel for free electrons ($m{=}m_0$, with $m_0$ being free electron mass.), but not generically in band systems. The orbital moment term $B M_z$, which generally couples the two-band subspace (of interest) to all other bands, is absent in any simplified two-band model.
%Observe that  that $\calh$ is actually time-independent, which originates from a  that is unbroken by the field (aligned with the rotational axis). Consequently,

@@

We have argued in \s{sec:qtznrules} that the subleading corrections ($\lambda_{1,2}$) to the quantization rule [cf.\ \q{eq:rule}] generically lift the spin-degeneracy of Landau levels -- the level splitting is approximately $\var_c(\lambda_1{-}\lambda_2)/2\pi$. Conversely, Landau levels are approximately spin-degenerate if $\lambda_1{=}\lambda_2$ (mod $2\pi$); this non-generic scenario may require fine-tuning of the Peierls-Onsager Hamiltonian.

In \s{sec:relatedegeneracies}, we will describe the precise relation between eigenvalue-degeneracies of  $\A$ and (near) degeneracies of Landau levels. We then ask how many tunable real parameters are necessary to fine-tune an eigenvalue-degeneracy of $\A$. Our answer is either $0,1,2$ or $3$, depending on the symmetry class of the orbit. Before providing this general answer in \s{sec:tenfold}, we will first develop physical intuition by studying two models [cf.\ \s{sec:singleparameterrashba} and \ref{sec:rotsymmbreakdown}] where only one tunable parameter is needed -- this can be the B field itself. Finally, the experimental implications of single- and double-parameter tuning of Landau-level crossings are discussed in \s{sec:qo}.

\subsection{Relating the eigenvalue-degeneracy of $\A$ to the spin-degeneracy of Landau levels}\la{sec:relatedegeneracies}

Suppose the propagator $\A$ were degenerate at $(\bar{E},\bar{B})$, i.e., $\lambda_+(\bar{E},\bar{B})=\lambda_-(\bar{E},\bar{B})$ mod $2\pi$. Generically, $(\bar{E},\bar{B})$ is not \emph{simultaneously} a solution of the quantization rule in \q{eq:rule}. That is to say, $\bar{E}$ is not generically the energy of a Landau level at field $\bar{B}$. To obtain a Landau level, we may tune either $E$ or $B$   away from $(\bar{E},\bar{B})$, and  utilize that  $\lambda_{\pm}$ are slowly-varying functions of $(E,B)$ compared to $l^2S(E)$:
\e{\left|l^2\p{S}{E}\right| \gg \left|\p{\lambda_a}{E}\right|,\, \left|S\right| \gg \left|\p{\lambda_a}{l^2}\right|;} 
these inequalities are guaranteed by the quasidegeneracy condition ($\delta S{\ll}S$) and the smoothness of $S(E)$, as we quantify in \app{sec:slowvariation}.  For fixed $\bar{B}$, we are therefore guaranteed to find two Landau levels (denoted $E_{\pm}$) in close proximity to $\bar{E}$, with a splitting of order:
\e{ \f{E_{+}-E_-}{\var_c}\bigg|_{\bar{B}} = \order\left( \f{\partial \lambda_a/\partial E}{l^2\partial  S/\partial E}  \right)\ll 1.\la{llquasideg}}
Alternatively, we may fix $\bar{E}$ and investigate how Landau levels disperse as a function of $B$; the analog of \q{llquasideg}  is the existence of two Landau levels (denoted $l^2_{\pm}$) in close proximity to $\bar{l}^2$, with a splitting of order:
\e{  \f{l^2_{+}-l^2_-}{T_{\sma{1/B}}}\bigg|_{\bar{E}} = O\left( \f{\partial \lambda_a/\partial l^2}{S} \right)\ll 1,\as T_{\sma{1/B}}:=\f{2\pi}{S(E)}.\la{llquasidegB} }
$\partial \lambda_a/\partial l^2$ has been estimated in \s{sec:slowvariation}, and
$T_{\sma{1/B}}$ above is the period of quantum oscillations (in the absence of spin-orbit coupling). To recapitulate, for every exact degeneracy of $\cala$, there must exist, in close proximity, a pair of Landau levels which are exactly degenerate, or nearly degenerate in the sense of \qq{llquasideg}{llquasidegB}. Such Landau levels will be called \textit{quasidegenerate}.

@@

In the absence of symmetry,  eigenvalue-degeneracies of the unitary propagator [cf. \q{eq:prop}] are attainable by tuning three real parameters. Analogously, it is well known (as the non-crossing rule in quantum theory[von Neumann,wigner]) that eigenvalue-degeneracies of complex, finite-dimensional matrix Hamiltonians are also attainable by tuning three real parameters.
 To prove the non-crossing rule for two-by-two unitaries, we begin by decomposing any such matrix as $\A{=}e^{i\phi}{\cal \bar{A}}$, such that  ${\cal \bar{A}}{\in}\text{SU}(2)$ being equal to $\pm I$ is a necessary and sufficient condition for an eigenvalue-degeneracy of $\A$. Suppose ${\cal \bar{A}}$ is parametrized by $d$ independent real numbers  $\{R_1,\ldots,R_{d}\}{:}{=}\bR $ which are tunable in experiments or on paper. Then ${\cal \bar{A}}(\bR){=}{\pm}I$ defines a set of manifolds in $\bR$-space which are henceforth referred to as degeneracy manifolds. The codimension of the degeneracy manifold  is defined as $d$ minus the dimension of said manifold; intuitively, this codimension is the number of linearly-independent parameters needed to fine-tune a degeneracy. Here and henceforth, we adopt the shorthand that real parameters will just be referred to as parameters, and \textit{codimension} always refers to a codimension of a degeneracy manifold. In the canonical parametrization of $\text{SU}(2)$ over the three-sphere $S^3$: 
\e{{\cal \bar{A}}{=}\matrixtwo{R_1{+}iR_2}{R_3{+}iR_4}{{-}R_3{+}iR_4}{R_1{-}iR_2},\as \sum_{j=1}^4R_j^2{=}1,  \la{s3}}
${\cal \bar{A}}{=}{+}I$ (resp.\ ${=}{-}I$) lie on a point which we refer to as the north pole (resp.\ south pole); both points have codimension three in $S^3$. If special unitaries were uniformly distributed over $S^3$, then encountering a spectral degeneracy for a randomly-selected special unitary would be an exceedingly rare event -- three parameters need be tuned to find either pole on $S^3$,  i.e., the codimension of degeneracy manifold is three. If only one  or two  parameters (in $\{R_1,\ldots,R_4\}$)  can be tuned, then in general we should not expect a degeneracy.

@@

It is well-known that symmetry may reduce codimension. The conceptually-simplest example is the approximate symmetry of band conservation, which has been described in \s{sec:qtznrules}. The salient fact to recall is that  $\cala$ becomes diagonal ($R_3{=}R_4{=}0$) in the basis of energy bands. The phase of $R_1{+}iR_2$ is then the single parameter  to tune $\cala{=}{\pm}I$, i.e., the codimension is unity.
From \q{phaseindependentorbit} we deduce that phase[$R_1{\pm}iR_2]{=}{\pm} l^2\delta S/2$ plus corrections which are field-independent, hence degeneracies of $\cala$ occur with a periodicity (in $l^2$) of $2\pi/\delta S$. This periodicity  has a dual interpretation  in real space -- as a periodic increment of a single flux quantum for  the differential magnetic flux (between the two spin-split orbits).

We may ask if these weak-field degeneracies (based on approximate band-conservation symmetry) persist in the strong-field regime where the adiabatic approximation breaks down. If there is no exact symmetry that persists at finite field, we have shown in the above demonstration [cf.\ \q{s3}] that the codimension is three, hence we should not expect robust degeneracies by tuning $B$ alone. Our next tasks are to identify the exact symmetries which persist in the presence of a finite magnetic field, and to identify which of the exact symmetries result in a reduced codimension.

\red{[Needs to introduce the symmetries more gently]} Since the magnetic field is odd under time reversal, and behaves as a pseudovector under spatial transformations, we deduce that exact symmetries (denoted $g$) are only of two types: either (i) $g$  preserves both the direction of time and the orientation of cyclotron orbits (in the 2D $\bk$-space perpendicular to $B$), or (ii) $g$ inverts both said objects. (i) is exemplified by a rotational symmetry with the rotational axis aligned parallel to the field, and (ii) by the composition  of time reversal $T$ and a spatial reflection $\mathfrak{r}_x$ (with reflection axis perpendicular to $B$).

One of our main results is that  rotational symmetry (of both discrete and continuous types) enforces unit codimension. This will be demonstrated by two  case studies in \s{sec:singleparameterrashba} and \ref{sec:rotsymmbreakdown}; the first study describes a Rashba-Dresselhaus 2DEG where the adiabatic approximation breaks down at all points along the semiclassical orbit; in the second study, the adiabatic approximation breaks down only in the vicinity of type-II Dirac points \textbf{[might mention angular oscillations]}.

In comparison, the composed symmetry $T\mathfrak{r}_x$ enforces a codimension of two. This result, in addition to a comprehensive codimension analysis of all relevant symmetry classes, is presented in \s{sec:tenfold}.

@@

\subsection{Rashba-Dresselhaus 2DEG}\la{sec:singleparameterrashba}

We return to the Rashba model absent Dresselhaus spin-orbit coupling and in-plane Zeeman field.  Owing to a continuous rotational symmetry (denoted $\mathfrak{c}_{\infty}$), $\calh(\bk)$ is a constant along the circular zeroth-order orbit [cf.\ \q{rashbaeffham}], and the propagator simplifies to $\A{=}$exp$[{-}i\calh T_c/\hbar]$. The latter is expressible in the exponential form  $\A =\exp[-i2\pi(r_1\tau_x{+}r_2\tau_z)]$, with $(r_1,r_2){=}(g_s'/4-1/2,-\alpha k_{\red{F}}ml^2\red{+1/2})\in\R^2$. In this two-dimensional parameter space, all degeneracies occur along circles of radius $\sqrt{r_1^2+r_2^2}{=}\red{j}/2$ with $\red{j}$ a nonnegative integer; half the circles ($\red{j}$ even) correspond to $\A=I$, and the other half to $\A=-I$. Except for $\red{j}=0$, all degeneracy manifolds have unit codimension, and can be tuned by varying $B$ alone (with fixed $g_s'$); these manifolds are analogous to line nodes (of Hamiltonian degeneracies) in 2D $\bk$-space -- a topic of recent interest. That tuning $\red{j}{=}0$ requires two parameters can be understood from an elementary result in quantum mechanics: taking $\alpha{=}0$ \textit{and} $m=m_0$, we recover the Zeeman-split Landau levels of a free electron gas without spin-orbit coupling; all levels (except the lowest) are spin-degenerate owing to the equality of the Zeeman and cyclotron energies. [LandauLifshitzstatphysI]   In this exactly-soluble model [cf.\ \q{eq:Rashba-exact}], all  degeneracies ($\red{j} \ge 0$) of $\A$ correspond to exact Landau-level degeneracies, rather than the near degeneracies described by \qq{llquasideg}{llquasidegB}.

@@

The Rashba model suffers from an oversimplification -- continuous rotational symmetry is not a symmetry of solids. That is to say, it does not belong in any (magnetic) space group. It is possible for $\mathfrak{c}_{\infty}$-symmetric models to approximate Hamiltonians with a discrete, order-$n$ rotational symmetry (denoted $\mathfrak{c}_N$) \red{[Let's use capital N in $\mathfrak{c}_N$ since N coincides with order]}.  To what extent are Landau-level degeneracies (of $\mathfrak{c}_{\infty}$-symmetric systems) robust when $\mathfrak{c}_{\infty}$ is perturbatively reduced to $\mathfrak{c}_N$? Our general answer is out of every $N$ degeneracies of $\A$, $N-1$ of them are robust. \textbf{[results for n=3,4,6 must be checked]}.

While full proof of the above claim is delegated to App. \ref{app:codimension}, let us illustrate the simplest example $N=2$ by introducing the Dresselhaus spin-orbit term ($\propto \beta$) into the Rashba model [cf.\ \q{hamRD}]. This has the desired effect of reducing $\mathfrak{c}_{\infty}$ to a discrete $\mathfrak{c}_2$ symmetry. Since $\A$ depends on $\alpha,\beta,l$ only through $\alpha l^2$ and $\beta l^2$, we consider the three-dimensional parameter space $(\alpha l^2,\beta l^2,g_s')\in \mathbb{R}^3$. If a  line node (in the $\beta{=}0$ plane) were perturbatively stable to finite $\beta$, then it would be embedded in a two-dimensional degeneracy hypersurface of $\mathbb{R}^3$; if the planar line node were unstable, this embedding does not happen, i.e., the planar line node is simply a line node in $\mathbb{R}^3$. Our numerical study shows that half the planar line nodes (corresponding to $\A{=}I$) are stable, while the other half ($\A{=}{-}I$) destabilizes. In $\mathbb{R}^3$, we then have pairs of line nodes and hypersurfaces enclosed by increasingly larger pairs of line nodes and hypersurfaces, as illustrated in Fig. \ref{fig:dgn}.

@@

Quasidegenerate orbits feature tunneling everywhere along the cyclotron orbit and therefore demand a quantization rule that incorporates both tunneling and single band evolution of wave packets. Sometimes, the coherent evolution of wave packets and tunneling can be decoupled in the quantization rule in weak field limit, as exemplified by magnetic breakdown in Rashba-Zeeman model in Sec. \ref{sec:inplanezeeman}. Let us further explore single-parameter tunability of Landau-level quasidegeneracies in this limit by introducing a rotational symmetric Hamiltonian with isolated breakdown junctions at several isolated $\bk$ points. Besides being a useful toy model, this Hamiltonian provides some insights of stability of Landau-level quasidegeneracies by perturbing from two otherwise independent orbits.

@@

\paragraph*{class II-A-1} In this class \textbf{[not just this class, also II-A-2]}, the full propagator can be constructed from $1/L$ of it, where $L$ is the smallest positive integer fulfilling $g^L\circ \bk_0=\bk_0$. Here, $\bk_0$ is the base point of $\A$. In this class, $L=N$. A good example in this class is $\mathfrak{c}_{Nz}$, $N$-fold rotation in z direction.\textbf{ [give the more general expression first before you perform your gauge transform]} Generally, by gauge transformation at $g^n\circ \bk_0$, $\breve{g}(g^n\circ \bk_0)$ can be transformed to identity for $0 \le n < N-1$ and the full propagator is
\e{
\A=\breve{g}(g^{N-1}\circ\bk_0)\prod_{i=0}^{N-1} \A_{1/N}
}
where 
\e{
\breve{g}(g^{N-1}\circ\bk_0)=\prod_{n=0}^{N-1} \breve{g}(g^n\circ\bk_0)
}
is a constant phase factor determined by the group multiplication rule. For example if $g$ is an $N$-fold rotation,  the phase factor is $+1$ (resp.\ $-1$) for an integer-spin (resp.\ half-integer-spin) representation. Therefore, the central object is
$\prod_{i=0}^{N-1} \A_{1/N}$. We take $N=2$ as an example. Using the parametrization in Eq. (\ref{eq:paramA1/N}),
\e{
\A_{1/N}^2=e^{2i\theta}\matrixtwo{\alpha^2-|\beta|^2}{\beta(\alpha+\alpha^*)}{-\beta^*(\alpha+\alpha^*)}{\alpha^{*2}-|\beta|^2}.
}
For $\A_{1/N}^2 \propto I$, the sufficient and necessary condition is $\text{Re}(\alpha)=0,\pm 1$, where tuning $\text{Re}(\alpha)=0$ requires only one parameter while tuning $\text{Re}(\alpha)=\pm 1$ requires three parameters.

>>>>

For simplicity, we assume that the Pauli matrices in \q{eq:Rashba-Hamiltonian} correspond to spin: $s_i{=}\hbar \sigma_i/2$. Let us evaluate the effective Hamiltonian $\calh(\bk)$  along the cyclotron trajectory $\frako_0$ with radius $k$. In the  energy basis of the Rashba Hamiltonian $H_R$ [cf.\ \q{eq:Rashba-Hamiltonian}], 
\e{\calh(\bk)=-\alpha k\tau_z{+}\f{g_s}{2}\mu_{\sma{B}}B\tau_x \red{-} \f{eB\hbar}{2m} (\tau_x-I) \epsilon_{\alpha\beta} k_{\red{\alpha}}\nabla_{\red{\beta}}\theta,\la{rashbaeffham}}
where $\tau_i$ are a new set of Pauli matrices with $\tau_z{=}{+} 1$ and ${-}1$ corresponding to the outer and inner orbits, respectively. $e^{i\theta_{\bk}}{=}(ik_x{-}k_y)/k$ is the angle of the in-plane spin-orbit field $({-}\alpha k_y,\alpha k_x,0)$, which winds once over $\frako_0$; the spin of our basis vectors likewise winds, in parallel (or anti-parallel) alignment with this field (see Fig. (\ref{fig:orbits})(b)). Terms proportional to the rate of spin rotation  ($\nabk \theta$) originate from the  non-abelian Berry connection. The first term of  $\calh$ corresponds to the spin-orbit-induced splitting of the energy-degenerate bands;
the remaining terms may be interpreted as B-induced deformations of the bands.   It is well-known that the Zeeman term tends to polarize spin in the direction ($\vec{z}$) of $\bB$; remarkably, the non-abelian Berry term   also has a polarizing effect, but with opposite sign. These two effects completely cancel for free electrons ($m{=}m_0$, with $m_0$ being free electron mass.), but not generically in band systems. The orbital moment term $B M_z$, which generally couples the two-band subspace (of interest) to all other bands, is absent in any simplified two-band model.
%Observe that  that $\calh$ is actually time-independent, which originates from a  that is unbroken by the field (aligned with the rotational axis). Consequently,

@@

We have argued in \s{sec:qtznrules} that the subleading corrections ($\lambda_{1,2}$) to the quantization rule [cf.\ \q{eq:rule}] generically lift the spin-degeneracy of Landau levels -- the level splitting is approximately $\var_c(\lambda_1{-}\lambda_2)/2\pi$. Conversely, Landau levels are approximately spin-degenerate if $\lambda_1{=}\lambda_2$ (mod $2\pi$) -- this degeneracy condition for the eigenvalues of $\A$ may require fine-tuning of the Peierls-Onsager Hamiltonian.

In \s{sec:relatedegeneracies}, we will describe the precise relation between eigenvalue-degeneracies of  $\A$ and (near) degeneracies of Landau levels. We then ask how many tunable real parameters are needed to attain an eigenvalue-degeneracy of $\A$. Our answer is  $3$ in the absence of symmetry, as discussed in \s{sec:introducecodimension}; the answer may reduce to $0,1$ or $2$ depending on the symmetry class of the orbit, as will be explained in \s{sec:tenfold}. Before this general symmetry classification,  we will first develop physical intuition by studying two rotationally-symmetric models [cf.\ \s{sec:singleparameterrashba} and \ref{sec:rotsymmbreakdown}] where only one tunable parameter is needed -- this can be the $B$ field itself. Finally, the experimental implications of single- and double-parameter tuning of Landau-level degeneracies are discussed in \s{sec:qo}.

\subsection{Relating the eigenvalue-degeneracy of $\A$ to the quasidegeneracy of Landau levels}\la{sec:relatedegeneracies}

Suppose the propagator $\A$ were degenerate at $(\bar{E},\bar{B})$, i.e., $\lambda_1(\bar{E},\bar{B})=\lambda_2(\bar{E},\bar{B})$ mod $2\pi$. Generically, $(\bar{E},\bar{B})$ is not \emph{simultaneously} a solution of the quantization rule in \q{eq:rule}. That is to say, $\bar{E}$ is not generically the energy of a Landau level at field $\bar{B}$. To obtain a Landau level, we may tune either $E$ or $B$   away from $(\bar{E},\bar{B})$, and  utilize that  $\lambda_{\pm}$ are slowly-varying functions of $(E,B)$ compared to $l^2S(E)$:
\e{l^2\left|\p{S}{E}\right| \gg \left|\p{\lambda_a}{E}\right|,\, \left|S\right| \gg \left|\p{\lambda_a}{l^2}\right|;} 
these inequalities are guaranteed by the quasidegeneracy condition ($\delta S{\ll}S$) and the smoothness of $S(E)$, as shown in \app{sec:slowvariation}.  For fixed $\bar{B}$, we are therefore guaranteed to find a number (of order one) of paired Landau levels  in close proximity to $\bar{E}$, with each pair having an energetic splitting of order:
\e{ \bigg|\f{E_{1}-E_2}{\var_c}\bigg|_{B=\bar{B}} = \order\left( \f{\partial \lambda_a/\partial E}{\bar{l}^2\partial  S/\partial E}  \right)\ll 1.\la{llquasideg}}
Alternatively, we may fix $\bar{E}$ and investigate how Landau levels disperse as a function of $B$; the analog of \q{llquasideg}  is the existence of paired Landau levels in close proximity to $\bar{l}^2$, with a splitting of order:
\e{  \bigg|\f{l^2_{+}-l^2_-}{T_{{l^2}}}\bigg|_{E=\bar{E}} = O\left( \f{1}{S}\p{\lambda_a}{l^2} \right)\ll 1,\;\; T_{{l^2}}:=\f{2\pi}{S(\bar{E})}.\la{llquasidegB} }
$\partial \lambda_a/\partial l^2$ is estimated in \app{sec:slowvariation}, and
$T_{l^2}$ above is the period of quantum oscillations (in the absence of spin-orbit coupling). To recapitulate, for every exact degeneracy of $\cala$, there must exist, in close proximity, a number of paired Landau levels which are exactly degenerate, or nearly degenerate in the sense of \qq{llquasideg}{llquasidegB}. Each such pair of Landau levels will be called \textit{quasidegenerate}.

@@

\subsection{Codimension analysis in the absence of symmetry}\la{sec:introducecodimension}

In the absence of symmetry,  eigenvalue-degeneracies of the unitary propagator $\A$ [cf. \q{eq:prop}] are attainable by tuning three real parameters. Analogously, it is well known (as the non-crossing rule in quantum theory[von Neumann,wigner]) that eigenvalue-degeneracies of complex, finite-dimensional matrix Hamiltonians are also attainable by tuning three real parameters.
 To prove the non-crossing rule for two-by-two unitaries, we begin by decomposing any such matrix as $\A{=}e^{i\phi}{\cal \bar{A}}$, such that  ${\cal \bar{A}}{\in}\text{SU}(2)$ being equal to $\pm I$ is a necessary and sufficient condition for an eigenvalue-degeneracy of $\A$. In the canonical parametrization of $\text{SU}(2)$ over the three-sphere $S^3$: 
\e{{\cal \bar{A}}{=}\matrixtwo{R_1{+}iR_2}{R_3{+}iR_4}{{-}R_3{+}iR_4}{R_1{-}iR_2},\;\; \sum_{j=1}^4R_j^2{=}1,\;\; R_j\in \mathbb{R},  \la{s3}}
${\cal \bar{A}}{=}{\pm}I$ lie on distinct points of $S^3$, and are attainable by tuning three real parameters: $R_3{=}R_4{=}0$, and the phase of $R_1{+}iR_2$ must equal $0$ or $\pi$. The conclusion that three parameters need be tuned remains valid for any  non-singular\footnote{By `non-singular parametrization', we mean a set of parameters related to those in \q{s3} by a non-singular coordinate transformation.} parametrization of SU$(2)$.

%If special unitaries were uniformly distributed over $S^3$, then encountering a spectral degeneracy for a randomly-selected special unitary would be an exceedingly rare event -- three parameters need be tuned to find either pole on $S^3$,  i.e., the codimension of degeneracy manifold is three. If only one  or two  parameters (in $\{R_1,\ldots,R_4\}$)  can be tuned, then in general we should not expect a degeneracy.

 To generalize the above discussion, let us parametrize ${\cal \bar{A}}$ by $d$ independent real numbers  $\{R_1,\ldots,R_{d}\}{:}{=}\bR $;  ${\cal \bar{A}}(\bR){=}{\pm}I$ defines a set of manifolds in $\bR$-space which we refer to as degeneracy manifolds; in the above example, these manifolds comprise two points on $S^3$. The codimension of the degeneracy manifold  is defined as $d$ minus the dimension of said manifold; alternatively stated, the codimension is the number of linearly-independent parameters needed to fine-tune a degeneracy. Here and henceforth, we adopt the shorthand that  \textit{parameters} will always refer to real parameters, and \textit{codimension} always refers to a codimension of a degeneracy manifold.

@@

It is well-known that symmetry may reduce codimension.[wigner,von neumann] The conceptually-simplest example is the approximate symmetry of band conservation, which exists in the adiabatic regime [cf.\ \s{sec:qtznrules}]. Recall from \q{phaseindependentorbit} that  $\cala$ becomes diagonal ($R_3{=}R_4{=}0$) in the basis of energy bands,
with phase[$R_1{\pm}iR_2]{=}{\pm} l^2\delta S/2$ plus field-independent corrections. Since this  phase alone  can be  tuned to attain $\cala{=}{\pm}I$, the codimension is unity.
As a function of $l^2$, we deduce that Landau-level quasi-crossings occur  with a periodicity of $2\pi/\delta S$. This periodicity  has a dual interpretation  in real space -- as a periodic increment of a single flux quantum for  the differential magnetic flux (between the two split orbits).

We may ask if these quasi-crossings (based on band-conservation symmetry) persist at stronger fields where adiabaticity breaks down. If no symmetry persists at finite field, then the codimension is three [cf.\ \q{s3}], hence Landau-level quasi-crossings should not generically be found by varying $B$ alone.

At finite field (assumed to be homogeneous in space and time), any remnant  symmetry (denoted $g$) must preserve the component of the magnetic moment in the direction of  $\bB$. The magnetic moment is odd under time reversal, and behaves as a pseudovector under spatial transformations.[sakurai] Therefore, either (i) $g$  preserves both the direction of time and the magnetic moment, or (ii) $g$ inverts both said objects. (i) is exemplified by a rotational symmetry with the rotational axis aligned parallel to the field, and (ii) by the composition  of time reversal $T$ and a spatial reflection $\mathfrak{r}_x$ (with reflection axis perpendicular to $\vec{B}$).

One of our main results is that  rotational symmetry (of both discrete and continuous types) results in degeneracy manifolds with unit codimension. This will be demonstrated by two  case studies: the Rashba 2DEG in \s{sec:singleparameterrashba}, and a model with II-Dirac points in \ref{sec:rotsymmbreakdown}. In contrast,  $T\mathfrak{r}_x$ results in manifolds with codimension two. The latter result, in addition to a comprehensive codimension analysis for all symmetry classes, is presented in \s{sec:tenfold}.

%The first study describes a Rashba-Dresselhaus 2DEG where the adiabatic approximation breaks down at all points along the semiclassical orbit; in the second study, the adiabatic approximation breaks down only in the vicinity of type-II Dirac points.

@@

\subsection{Unit codimension for the Rashba 2DEG}\la{sec:singleparameterrashba}

To exemplify the role of rotational symmetry in reducing codimension,
we return to the Rashba model [cf.\ \q{eq:Rashba-Hamiltonian}] absent Dresselhaus spin-orbit coupling and in-plane Zeeman field.  Owing to a continuous rotational symmetry (denoted $\mathfrak{c}_{\infty}$), $\calh(\bk)$ is a constant along the circular zeroth-order orbit [cf.\ \q{rashbaeffham}], and therefore the propagator simplifies to $\A{=}$exp$[{-}i\calh T_c/\hbar]$. The latter is expressible in the exponential form:
\e{\A =e^{-i2\pi(r_1\tau_x{+}r_2\tau_z)}, \;\;(r_1,r_2){=}\bigg(\f{g_{\sma{\text{eff}}}}{4}-\f1{2},\f1{2}-\f{\alpha k_{{E}}}{\var_c}\bigg).\notag} 
All degeneracies occur along circles of radius $\sqrt{r_1^2+r_2^2}{=}{j}/2$, with ${j}$ a nonnegative even (resp.\ positive odd) integer for  $\A{=}I$ (resp.\ $\A{=}{-}I$). In this exactly-soluble model [cf.\ \q{eq:Rashba-exact}], all  degeneracies (${j} {\ge} 0$) of $\A$ correspond to exact Landau-level degeneracies, rather than the near degeneracies described by \qq{llquasideg}{llquasidegB}.

That the ${j}{=}0$ manifold has codimension two can be understood from an elementary result in quantum mechanics: taking $\alpha{=}0$ \textit{and} $m{=}m_0$, we recover the Zeeman-split Landau levels of a free electron gas without spin-orbit coupling; all levels (except the lowest) are spin-degenerate owing to the equality of the Zeeman and cyclotron energies.[LandauLifshitzstatphysI]   All other manifolds ($j{>}0$) have unit codimension, and can be tuned by varying $B$ alone (with fixed $g_{\sma{\text{eff}}}$); these manifolds are analogous to line nodes (of Hamiltonian degeneracies) in 2D $\bk$-space -- a topic of recent interest.[citations]

@@

The Rashba model suffers from an oversimplification -- continuous rotational symmetry is not a symmetry of solids. That is to say, it does not belong in any crystallographic space group. It is possible for $\mathfrak{c}_{\infty}$-symmetric $\bk{\cdot}\bp$ models to approximate Bloch Hamiltonians (defined over the Brillouin torus) with a discrete, order-$N$ rotational symmetry (denoted $\mathfrak{c}_N$). To what extent are degeneracies of $\A$ (for $\mathfrak{c}_{\infty}$-symmetric systems) robust when $\mathfrak{c}_{\infty}$ is perturbatively reduced to $\mathfrak{c}_N$? For each of $N{=}2,3,4$ and $6$ (the allowed values in space groups), we find that $N{-}1$ out of every $N$ degeneracies (of $\A$) are robust, as proven in App. \ref{app:codimension}.

For illustration, let us reduce $\mathfrak{c}_{\infty}$ symmetry to  $\mathfrak{c}_2$  by adding the Dresselhaus spin-orbit term ($\propto \beta$) to the Rashba model [cf.\ \q{hamRD}].  Since $\A$ depends on $\alpha,\beta,l$ only through $\alpha l^2$ and $\beta l^2$, we consider the three-dimensional parameter space $(\alpha l^2,\beta l^2,g_{\sma{\text{eff}}})\in \mathbb{R}^3$. If a  line node (in the $\beta{=}0$ plane) were perturbatively stable to finite $\beta$, then it would be embedded in a two-dimensional degeneracy hypersurface of $\mathbb{R}^3$; if the planar line node were unstable, this embedding does not happen, i.e., the planar line node is simply a line node in $\mathbb{R}^3$. Our numerical study shows that half the planar line nodes (corresponding to $\A{=}I$) are stable, while the other half ($\A{=}{-}I$) destabilizes. In $\mathbb{R}^3$, we then have pairs of line nodes and hypersurfaces enclosed by increasingly larger pairs of line nodes and hypersurfaces, as illustrated in Fig. \ref{fig:dgn}.

@@

Quasidegenerate orbits feature tunneling everywhere along the cyclotron orbit and therefore demand a quantization rule that incorporates both tunneling and single band evolution of wave packets. Sometimes, the coherent evolution of wave packets and tunneling can be decoupled in the quantization rule in weak field limit, as exemplified by magnetic breakdown in Rashba-Zeeman model in Sec. \ref{sec:inplanezeeman}. Let us further explore single-parameter tunability of Landau-level quasidegeneracies in this limit by introducing a rotational symmetric Hamiltonian with isolated breakdown junctions at several isolated $\bk$ points. Besides being an interesting toy model, this Hamiltonian provides some insights of stability of Landau-level quasidegeneracies by perturbing from two otherwise independent orbits.

@@

\paragraph*{class II-A-1} In this class \textbf{[not just this class, also II-A-2]}, the full propagator can be constructed from $1/L$ of it, where $L$ is the smallest positive integer fulfilling $g^L\circ \bk_0=\bk_0$. Here, $\bk_0$ is the base point of $\A$. In this class, $L=N$. A good example in this class is $\mathfrak{c}_{Nz}$, $N$-fold rotation in z direction.\textbf{ [give the more general expression first before you perform your gauge transform]} Generally, by gauge transformation at $g^n\circ \bk_0$, $\breve{g}(g^n\circ \bk_0)$ can be transformed to identity for $0 \le n < N-1$ and the full propagator is
\e{
\A=\breve{g}(g^{N-1}\circ\bk_0)\prod_{i=0}^{N-1} \A_{1/N}
}
where 
\e{
\breve{g}(g^{N-1}\circ\bk_0)=\prod_{n=0}^{N-1} \breve{g}(g^n\circ\bk_0)
}
is a constant phase factor determined by the group multiplication rule. For example if $g$ is an $N$-fold rotation,  the phase factor is $+1$ (resp.\ $-1$) for an integer-spin (resp.\ half-integer-spin) representation. Therefore, the central object is
$\prod_{i=0}^{N-1} \A_{1/N}$. Using the parametrization in Eq. (\ref{eq:paramA1/N}), we diagonalize $A_{1/N}$ as 
\e{
A_{1/N} = e^{i\theta}U^{\dagger}\matrixtwo{e^{i\varphi}}{0}{0}{e^{-i\varphi}}U
}
with $\cos(\phi)=\text{Re}(\alpha)$. The sufficient and necessary condition for $A_{1/N}\propto I$ is $\text{Re}(\alpha)=\cos(n\pi/N)$, with $n\in\mathbb{Z}$. Tuning $\text{Re}(\alpha)=\cos(n\pi/N)$ requires only one parameter, except for $n=0$ and $n=N$, where three parameters are required.

In the Sec. \ref{sec:llquasideg}, we construct $\mathfrak{c}_N$ models by perturbing from $\mathfrak{c}_\infty$ Rashba model and by coupling two otherwise independent orbits by magnetic breakdown. In both constructions, $\varphi$ evolves monotonically with respect to $B$ before perturbation are turned on. $\varphi$ sweeps through $n\pi/N$ where degeneracies of $\A$ are expected. $e^{i\varphi}$ hits $\pm 1$ once in every $N$ degeneracies and therefore the other $n-1$ degenerates are stable, i.e., have codimension 1.
